Staged Abstract Interpreters

### 2019-03-25 paper/30_stageconc.tex
section 3

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to staged version -- we just need to let them works on staged data.

\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[(Value, Store)]
  def emit_compiled_clo(fun: (Rep[Value], Rep[Store]) => Rep[(Value, Store)],
                        λ: Lam, ρ: Rep[Env]): Rep[Value]

  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ <- get_store
    val res: Rep[(Value, Store)] = emit_ap_clo(fun, arg, σ)
    _ <- put_store(res._2)
  } yield res._1
\end{lstlisting}

@@

\begin{lstlisting}
  case class IRApClo(f: Exp[Value], arg: Exp[Value], σ: Exp[Store]) extends Def[(Value, Store)]
  override def emit_ap_clo(fun: Exp[Value], arg: Exp[Value], σ: Exp[Store]): Exp[(Value, Store)] =
    IRApClo(fun, arg, σ)
\end{lstlisting}

\begin{lstlisting}
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
    case IRApClo(f, arg, σ) =>
      emitValDef(sym, s"$\textdollar${quote(f)}.asInstanceOf[CompiledClo].f($\textdollar${quote(arg)}, $\textdollar${quote(σ)})")
    ...
  }
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the staging part, we share the same concrete type instantiation, i.e., trait
\texttt{Concrete}, but we reimplement the staged version for the concrete
operations in trait @RepConcInterpOps@. Meanwhile, it is also extended from
@LMSOps@ which provides necessary support for the staged version of primitive
types and data structures. To stage the interpreter, we first identify the
syntactic input that is known statically, and the interpreter returns a pair of
values and store, which both becomes dynamic values. Thus the staged stores
propagate the binding-time to environments, because they share the same address
component.

The first notable change is that the abstract type member @R@ is assigned to be
@Rep@, and accordingly the affected types such as @Env@ and @Store@ become
@Rep[Env]@ and @Rep[Store]@. For @num@, we use the @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage. The interesting
point is how we handle closures and function applications when staging is
involved. In the unstaged version, function @close@ is used to lift a literal
lambda term to a closure, which is a pair of the syntactic lambda term and the
enclosing environment. However, what we desire is a \textit{compiled} closure,
instead of a lifted @CloV@ value like what we did for @NumV@ --- in other words,
the syntactic term of the lambda expression should be eliminated and specialized
away. The specialization of the interpreter with respect to the body of the
lambda term proceeds under the assumptions that the argument and latest store
will be provided later.

We will see this is an important observation that enables us to specialize an
abstract interpreter in a modular way. At the end of function @close@, we create
a @CompiledClo@ object that contains a staged function, which all will be
emitted in the generated code.

\begin{lstlisting}
def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
  val Lam(x, e) = $\lambda$
  val f: Rep[(Value,Store)]=>Rep[(Value,Store)] = { 
    case (arg: Rep[Value],$\sigma$: Rep[Store]) =>
      val $\alpha$ = alloc($\sigma$, x)
      ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, arg)) 
  }
  CompiledClo(fun(f))
}
\end{lstlisting}

Next, @apply_closure@ takes three arguments: the function value, argument value
and the latest store, which are all of @Rep@ type. This means basically there
is nothing we can do at the current stage --- so we create a new node
@ApplyClosure@ that contains these arguments in the intermediate representation
graph, and in the code generation phase we can emit the code that performs the
application. The additional @reflectEffect@ is a function in LMS that handles
side effects.

\begin{lstlisting}
case class ApplyClosure(f: Rep[Value], arg: Rep[Value], $\sigma$: Rep[Store]) extends Def[(Value, Store)]
def apply_closure(ev: EvalFun)
  (f: Rep[Value], arg: Rep[Value], $\sigma$: Rep[Store]): Ans = {
    reflectEffect(ApplyClosure(f, arg, $\sigma$))
  }
\end{lstlisting}

When generating code for the next stage, values like @CompiledClo@ and @NumV@
are kept as they are because they are intended exist in the next stage; however,
we need to treat @ApplyClosure@ a little differently.

Concretely, as shown in the code above, when emitting code for @ApplyClosure@,
we just emit a function application where the function value is extracted from
the @CompiledClo@ object (we assume the input program is well-typed so that we
can safely cast it to @CompiledClo@). Argument and a store are the remaining rest
fields in @ApplyClosure@.

>>>>

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to staged version -- we just need to let them works on staged data.
As we mentioned before, we elaborate how to handle functions and applications in
detail. One of the desired goal is to eliminate the monadic layers through
staging, this becomes tricky for clsoures, because we need to actually compile
the lambda term instead of generate a @CloV@ value. To make this happen, when
closing the lambda term with the environment, we need to \textit{collpase} the
monads to normal values. The following function @close@ illustrates the idea:
@f@ is a current stage function but takes two next-stage values @v@ and
$\sigma$, inside of which, we eagerly collapse the monads to values by
@ev(e)(ρ_*)(σ_*)@, which has type @Rep[(Value, Store)]@.
Then we generate an intermediate representation of the compiled closure using
@emit_compiled_clo@. The intermediate representation stands for a next-stage
value, and contains the compiled function, which will be generated as a
next-stage function.

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)]): Rep[Value]
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ + (α + v)
        ev(e)(ρ_*)(σ_*)
    }; emit_compiled_clo(f)
  }
\end{lstlisting}

When applying a function in @ap_clo@, we generate an intermediate representation
for closure application through @emit_ap_clo@, assuming that the @rator@ is already an
IR of compiled closure, @rand@ is an arbitrary stage value, and $\sigma$ is the
latest store. @emit_ap_clo@ represents a next-stage value of type
@Rep[(Value, Store)]@, which is a collapsed monadic value from next stage, now
we need to reify the value and store back into the monad stack through
@put_store@ and @yield@.

\begin{lstlisting}
  def emit_ap_clo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]): Rep[(Value, Store)]
  def ap_clo(ev: EvalFun)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store
    val res: Rep[(Value, Store)] = emit_ap_clo(rator, rand, σ)
    _ <- put_store(res._2)
  } yield res._1
\end{lstlisting}

@@

The method @emit_compiled_clo@ and @emit_ap_clo@ mentioned above produces IR
node in LMS. We define these IR node as follow: \note{Def, quote are omitted}

\begin{lstlisting}
  case class IRCompiledClo(f: Rep[((Value,Store)) => (Value, Store)]) extends Rep[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Rep[(Value, Store)]
\end{lstlisting}

After constructing the IR graph, LMS provides an @emitNode@ to generate code for
each kind of IR node. To generate code for @IRApClo@, we know that @rator@ is a
compiled closure @IRCompiledClo@, which contains a next-stage function @f@, so
we just need to generate code that invokes @f@ with @rand@ and @σ@ as arguments
provided. Similarly, we directly generate a next-stage value @CompiledClo(f)@
for @IRCompiledClo@.

\begin{lstlisting}
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
    case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar$rand, $\textdollar$σ)")
    case IRCompiledClo(f) => emitValDef(sym, s"CompiledClo($\textdollar$f)")
    ...
  }
\end{lstlisting}

### 2019-03-25 paper/50_stageabs.tex
section 3

\subsection{Specialized Data Structures} \label{staged_ds}

@@

@@

In the literature of partial evaluation, \citeauthor{10.1007/3-540-61580-6_11}
provided guidelines on what to do and not to do when specializing a concrete
interpreter \cite{10.1007/3-540-61580-6_11}. We borrow these guidelines and
extend them to abstract interpreters. We discuss decisions we made to achieve
this and examine some alternatives.

@@

What we implemented is a big-step, compositional abstract interpreter, where
"compositional" means that every recursive call of our abstract interpreter is
applied to proper substructures of the current syntactic parameters
\cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as that the specialization
procedure terminates. Nevertheless, it is also possible to specialize a
small-step operational abstract semantics ---
\citeauthor{Johnson:2013:OAA:2500365.2500604} showed this in abstract
compilation \cite{Boucher:1996:ACN:647473.727587} style as one of their
optimizations of Abstract Abstract Machines
\cite{Johnson:2013:OAA:2500365.2500604}. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts. Another alternative approach for efficient
specialization is to write the abstract interpreter in a big-step, monadic style
\cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

Based on the assumption that LMS preserves the semantics during staging, we are
confident that the staged abstract interpreter does the same analysis compared
with the unstaged one. Moreover, the optimization done by staging does not
compromise any soundness.

>>>>

@@

\subsection{Optimizations} \label{staged_ds}

Our staging schema works well in theory, but would suffer from code explosion,
and possibly runtime GC overhead when analyzing (specializing) large programs.
We present several optimizations that largely mitigate these issues. Still,
these optimizations do not need to change the generic interpreter.

\paragraph{Specialized Data Structures}

\paragraph{Selective Caching} It is observed that the two-fold caching is
wrapped for recursive call of our abstract interpreter. But this is even more
expensive when generating code for atomic expressions such as literal numbers or
variables. \todo{caching prevents nontermination}

\paragraph{Partially-static Data}

For example, to fold a singleton list, e.g., @List(x).foldLeft(init)(f)@ where
@x@ and @init@ are staged values, a naive code generator would faithfully
apply @foldLeft@ to the list with an anonymous function. But we can also utilize
the algebraic property of @foldLeft@ to generate cheaper code:
@List(x).foldLeft(init)(f) = f(init, x)@.
Since the function @f@ is known at current stage, we completely eliminate the
fold operation and function application. We apply several rewritings enabled by
partial-static data structures, such as @List@ and @Map@, which greatly reduce
the size of residual programs.

\paragraph{Heterogeneous Staging} The generated program is in A-Normal form,

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters. In this section, we discuss several decisions we made to
achieve this and examine some alternatives.

\paragraph{Are Monads Necessary?} No. One can always inline the monads and
obtain an abstract interpreter in continuation-passing style, or simply use
explicit side-effects such mutation to implement the artifact with the same
abstract semantics. In either cases, we can still apply the staging schema to
abstract interpreter. As an evidence, we provide an staged abstract interpreter
written in direct-style in the accompanying artifact.

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where "compositional" means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as the termination of
specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} --
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizating Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts.

@@

As one of the merit of our approach, the staging does not compromise any
soundness, because we only change the underlying monads and data structures to
the staged version, not any part of the abstract semantics. Based on the
assumption that LMS preserves the equivalence during staging, we are confident
that the staged abstract interpreter does the same analysis as the unstaged one.
